---
title: "Areal 2"
author: "AyouB"
date: "14/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Areal part

# Loading packages

```{r }
pacman::p_load(tidyverse,
               INLA,
               sp,
               sf,
               rgdal,
               rgeos,
               tmap,
               tmaptools,
               readxl,
               RColorBrewer,
               spdep,
               raster,
               patchwork)
```

## multi thread option 

```{r}
inla.setOption(num.threads=1)
```


# Help Functions


```{r }
logit <- function(x){log(x/(1-x))}
expit <- function(x) {exp(x)/(1+exp(x))}
```

# Data

```{r}
# Import Data
data = read_xlsx("data.xlsx", sheet = "Sheet1")
data$ID <- 1:48

# Getting shapefiles
dz_mapw <- getData("GADM", country ="DZ", level = "1")
dz_map <- getData("GADM", country ="DZ", level = "0")

dz2 <- st_as_sf(dz_mapw)

# cleaning wilayas names
Wilaya <- dz2$NAME_1 %>% 
                lapply(gsub, pattern="ï", replacement = "i") %>% 
                unlist()
dz2$Wilaya <- Wilaya

# Merging data with shapefiles
full <- dz2 %>% full_join(data, by = "Wilaya")
full$CIG = full$CIG/1000

# Convert to spatial polygon again !
full_poly <- as_Spatial(full)
```


# Some Exploratory data analysis 

```{r}
full_poly@data %>%
  ggplot(aes(x = Y)) +
  geom_histogram(bins = 10, fill = "blue") +
  geom_vline(xintercept = mean(data$Y), col ="red") +
  labs(
       x="Mortalité", y="Fréquence") +
  theme_bw()

full_poly@data %>%
  ggplot(aes(y = Y)) +
  geom_boxplot(fill = "lightblue") +
  geom_hline(yintercept = mean(data$Y), col ="red") +
  labs(y="Mortalité") + 
  theme_bw()

full_poly@data %>%
  ggplot(aes(x = E)) +
  geom_histogram(bins = 10, fill = "blue") +
  geom_vline(xintercept = mean(data$E), col ="red") +
  labs(
       x="Mortalité attendue", y="Fréquence") +
  theme_bw()

full_poly@data %>%
  ggplot(aes(y = Y)) +
  geom_boxplot(fill = "lightblue") +
  geom_hline(yintercept = mean(data$E), col ="red") +
  labs(y="Mortalité espérée") +
  theme_bw()


full_poly@data %>%
  ggplot(aes(x = CIG)) +
  geom_histogram(bins = 10, fill = "blue") +
  geom_vline(xintercept = mean(data$CIG)/1000, col ="red") +
  labs(
    x="Nombre de cigarettes (100 Millions)", y="Fréquence") +
  theme_bw()

full_poly@data %>%
  ggplot(aes(y = CIG)) +
  geom_boxplot(fill = "lightblue") +
  geom_hline(yintercept = mean(data$CIG)/1000, col ="red") +
  labs(y="Nombre de cigarettes (100 Millions)") + 
  theme_bw()
```


## Neighboorhood analysis 


# Creation of neighboors

```{r}
temp <- poly2nb(full_poly)
nb2INLA("ALG.graph", temp)
ALG.adj <- paste(getwd(),"/ALG.graph",sep="")

wmat <- nb2listw(temp, style ="C")

plot(dz_mapw, border = "grey60")
plot(wmat, coordinates(dz_mapw), pch = 19, cex = 0.6, add = TRUE, col ="red")
```

# Moran Test 

```{r}
moran.test(full_poly@data$Y, listw = wmat, alternative = "less")
moran.test(full_poly@data$Y, listw = wmat, alternative = "greater")
moran.test(full_poly@data$Y, listw = wmat, alternative = "two.sided")



moran.plot(full_poly@data$Y, listw = wmat, zero.policy=T, pch=16, 
           col="black",cex=.5, quiet=F,
           labels=as.character(full_poly@data$Wilaya),
           xlab="Mortalité par cancer du poumon", 
           ylab="Mortalité par cancer du poumon (retard spatial)", 
           main="Nuage de points de Moran")
```

# Moran Test MC

```{r}
set.seed(1234)
mc = moran.mc(full_poly@data$Y, listw = wmat, nsim = 999, alternative = "greater")
mc

plot(mc, main="", las=1, xlab = "Mortalité")

s1 <- sp.mantel.mc(full_poly@data$Y, listw = wmat, nsim = 999, type ="moran", alternative = "greater")
s1
plot(s1)
```


# Geary Test 

```{r}
set.seed(1234)
geary.test(full_poly@data$Y, listw = wmat, zero.policy=T, alternative = "greater")
g.mc <- geary.mc(full_poly@data$Y, listw = wmat, nsim = 999, alternative = "greater")
g.mc
plot(g.mc, main="Densité des Permutations", las=1, xlab = "Mortalité", ylab = "Densité")

s2 <- sp.mantel.mc(full_poly@data$Y, listw = wmat, nsim = 999, type ="geary", alternative = "less")
s2
plot(s2)

```


# Local Spatial Heteroscedastisticity

```{r}
LOSH(full_poly@data$Y, listw = wmat)
l<- LOSH.cs(full_poly@data$Y, listw = wmat)
l
hist(l[,"Hi"])

lmc<- LOSH.mc(full_poly@data$Y, listw = wmat, 2 ,999)
plot(l[,"Pr()"], lmc[,"Pr()"])
```


# Local Moran 

```{r}
loc_moran <- localmoran(full_poly@data$Y, listw = wmat, zero.policy=T)
loc_moran

hist(loc_moran[,5])
```


# Save Local Moran (if needed)

```{r}
l <- as.data.frame(as.matrix(loc_moran))
# xlsx::write.xlsx(l, "local.xlsx")
```


# Plot Local Moran (you need the charge the function from the attached file)

```{r}
plot.local.moran(full_poly, variable.name = "Y", weights = wmat,sig =0.1,
                 local.moran = loc_moran, legend.location = "topright")
```

# Getis-Ord Test 

```{r}
globalG.test(full_poly@data$Y, listw = wmat, zero.policy=T, alternative = "greater")

getis <- localG(full_poly@data$Y, listw = wmat, zero.policy=T)
getis
```

# Some plots 

```{r}
full_poly@data$loc_moran <- abs(loc_moran[,4])
full_poly@data$loc_moran2 <- loc_moran[,1]
full_poly@data$getis <- as.numeric(getis)


full$loc_moran <- abs(loc_moran[,4])
full$loc_moran1 <- loc_moran[,1]
full$getis <- as.numeric(getis)

ggplot(full) + geom_sf(aes(fill = loc_moran)) +
  scale_fill_gradient2(midpoint = 1, low = "yellow", mid = "white", 
                       high = "red", limits = c(0, 2)) +
  theme_bw()

ggplot(full) + geom_sf(aes(fill = loc_moran1)) +
  scale_fill_gradient2(midpoint = 0, low = "blue", mid = "white", 
                       high = "red", limits = c(-0.5, 0.66)) +
  theme_bw()

ggplot(full) + geom_sf(aes(fill = getis)) +
  scale_fill_gradient2(midpoint = 1 , low = "green", mid = "white", 
                       high = "red", limits = c(-1.86, 3.4)) +
  theme_bw()

tm_shape(full_poly) + 
  tm_borders(col="black") +
  tm_fill(col = "loc_moran", title = "I de Moran locale (|z| scores)")

tm_shape(full_poly) + 
  tm_borders(col="black") +
  tm_fill(col = "loc_moran2", midpoint = NA, 
          palette = get_brewer_pal("-RdYlGn", n = 7), 
          title = "I de Moran locale (Valeurs)")

tm_shape(full_poly) + 
  tm_borders(col="black") +
  tm_fill(col = "getis", midpoint = NA, 
          palette = get_brewer_pal("-RdYlGn", n = 7), 
          title = "Local Getis-Ord G (Values)")
```

## Areal Modeling 

# Formulas 

```{r}
# the formulas changes with the model
formula.cov <- Y ~ 1+ + CIG + f(ID, model = "bym", graph = ALG.adj, 
                              hyper = list(prec.unstruct = list(prior="loggamma",param=c(0.1,0.01)), 
                                          prec.spatial = list(prior="loggamma",param=c(0.1,0.001))))

```

# The INLA Model

```{r}
mod.cov <- inla(formula.cov, family = "poisson", data = full_poly@data, E = E, verbose = F, 
                control.compute = list(cpo = TRUE, dic = TRUE, graph = TRUE, waic =TRUE),
                control.predictor=list(compute=TRUE))
```

# Model Summary 

```{r}
mod.cov$summary.fixed

mod.cov$summary.hyperpar
```

# Spatial Fraction 

```{r}
m <- mod.cov$marginals.random$ID[1:48]
zeta.cov <- lapply(m,function(x)inla.emarginal(exp,x))

a=0
a1 = 1

inlaprob.cov<-lapply(mod.cov$marginals.random$ID[1:48], function(X){
  1-inla.pmarginal(a, X)
})


inlaprob.cov2<-lapply(mod.cov$marginals.fitted.values, function(X){
  1-inla.pmarginal(a1, X)
})


m<-mod.cov$marginals.random$ID

mat.marg<-matrix(NA, nrow=48, ncol=1000)

for (i in 1:48){
  u<-m[[48+i]]
  s<-inla.rmarginal(1000, u)
  mat.marg[i,]<-s}

var.RRspatial<-mean(apply(mat.marg, 2, sd))^2
var.RRhet<-inla.emarginal(function(x) 1/x,
                          mod.cov$marginals.hyper$"Precision for ID (iid component)")
ss <- var.RRspatial/(var.RRspatial+var.RRhet)
ss
```

# Model evaluation

```{r}
mod.cov$dic$dic
mod.cov$dic$p.eff
mod.cov$waic$waic
mod.cov$waic$p.eff
mod.cov$mlik[1]

-sum(log(mod.cov$cpo$cpo))
```


# defaillance probability

```{r}
fail <- mod.cov$cpo$failure
fail 
sum(mod.cov$cpo$failure >0)

xlsx::write.xlsx(fail, "failure.xlsx")
```


# CPO plot 

```{r}
cpod <- data.frame(ID = 1:48,
                   wilayas = full$Wilaya,
                   CPO = mod.cov$cpo$cpo,
                   CPO2 = mod2$cpo$cpo
                   )

cpod %>%
  ggplot(aes(x = ID, y = CPO)) + 
  geom_point(col = "blue", alpha = 0.5) +
  geom_text(aes(label = wilayas)) + 
  geom_hline(yintercept = 0.01, col = "red") +
  theme_bw()

cpod %>%
  ggplot(aes(x = ID, y = CPO2)) + 
  geom_point(col = "blue", alpha = 0.5) +
  geom_text(aes(label = wilayas)) + 
  geom_hline(yintercept = 0.01, col = "red") + 
  labs(y ="CPO") +
  theme_bw()

```

# PIT plot 

```{r}
pit <- mod2$cpo$pit
uniquant <- (1:48)/(48+1)
plot(uniquant, sort(pit), xlab="quantiles uniformes",
     ylab="valeurs PIT ordonn?es")
abline(0,1)


plot(logit(uniquant), logit(sort(pit)), xlab="uniform quantiles", 
     ylab="Sorted PIT values", main="Logit scale")
abline(0,1)

which(abs(logit(pit)) > 3)


data.frame(pit = sort(pit), uni = uniquant) %>%
  ggplot(aes(x= uni, y = pit)) + 
  geom_point()+
  geom_abline(intercept = 0, slope = 1, col = "red") + 
  labs(x= "quantiles uniformes", y = "valeurs PIT ordonnées") +
  theme_bw()

data.frame(pit = logit(sort(pit)), uni = logit(uniquant)) %>%
  ggplot(aes(x= uni, y = pit)) + 
  geom_point()+
  geom_abline(intercept = 0, slope = 1, col = "red") +
  labs(x= "quantiles uniformes", y = "valeurs PIT ordonnées") +
  ggtitle("échelle logarithmique")+
  theme_bw()
```

# Model correction(takes time (variant with the number of CPO values))

```{r}
mod2 <- inla.cpo(mod.cov)
```

# SMR and RR

```{r}
full$smr <- full$Y/full$E
full$RR <- mod2$summary.fitted.values[, "mean"]
full$LL <- mod2$summary.fitted.values[, "0.025quant"]
full$UL <- mod2$summary.fitted.values[, "0.975quant"]

summary(full[, c("RR", "LL", "UL")])

full$smr
full$smr-full$RR
(full$smr-full$RR)^2
sum((full$smr-full$RR)^2)
sum((full$smr-full$RR)^2)/48
sqrt(sum((full$smr-full$RR)^2)/48)
```


# Plots SMR and RR

```{r}
gRR <- ggplot(full) + geom_sf(aes(fill = RR)) +
  scale_fill_gradient2(midpoint = 1, low = "blue", mid = "white", high = "red", limits = c(0.8, 1.3)) +
  theme_bw()

gLL <- ggplot(full) + geom_sf(aes(fill = LL)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red") +
  theme_bw()

gUL <- ggplot(full) + geom_sf(aes(fill = UL)) +
  scale_fill_gradient2(midpoint = 1, low = "blue", mid = "white", high = "red") +
  theme_bw()

gLL + gRR + gUL

p1 <- ggplot(full) + geom_sf(aes(fill = smr)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", 
    high = "red", limits = c(0.7,1.7)) + 
  theme_bw() +
  labs(fill = "SMR")

p2 <- ggplot(full) + geom_sf(aes(fill = RR)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", 
    high = "red", limits = c(0.8,1.3)) + 
  theme_bw() +
  labs(fill = "RR")

p1 + p2
```

# boxplot SMR

```{r}
full %>%
  ggplot(aes(y = smr)) +
  geom_boxplot(fill = "lightblue") +
  geom_hline(yintercept = mean(full$smr), col ="red") +
  labs(y="Taux de Mortalité Standardisé (SMR)") + 
  theme_bw()

full %>%
  ggplot(aes(y = RR)) +
  geom_boxplot(fill = "lightblue") +
  geom_hline(yintercept = mean(full$RR), col ="red") +
  labs(y="Risque Relatif (RR)") + 
  theme_bw()
```


# SMR and Probability of excedence 

```{r}
full <- full %>% mutate(SMR.zeta = unlist(zeta.cov),
                        pp.zeta = unlist(inlaprob.cov),
                        pp.eta = unlist(inlaprob.cov2))


ggplot(full) + geom_sf(aes(fill = SMR.zeta)) +
  scale_fill_gradient2(midpoint = 1.2, low = "blue", mid = "white", 
                       high = "red", limits = c(0.8, 1.3)) +
  theme_bw() +
  labs(fill ="SMR")

ggplot(full) + geom_sf(aes(fill = pp.zeta)) +
  scale_fill_gradient2(midpoint = 0.5, low = "green", mid = "white", 
                       high = "red", limits = c(0, 1)) +
  theme_bw() +
  labs(fill ="Zeta")

ggplot(full) + geom_sf(aes(fill = pp.eta)) +
  scale_fill_gradient2(midpoint = 0.5, low = "green", mid = "white", 
                       high = "red", limits = c(0, 1)) +
  theme_bw() + 
  labs(fill ="Probabilité a posteriori")
```


# some plot 

```{r}
full <- 
  full %>% mutate(re = mod.cov$summary.random$ID[1:48, "mean"])

ggplot(full) + 
  geom_sf(aes(fill = re)) +
  scale_fill_gradient2(
    midpoint = 0, low = "blue", mid = "white", high = "red"
  ) +
  theme_bw()
```




# BYM with PC prior 

```{r}
formula.cov <- Y ~ 1+ +CIG+f(ID, model = "bym2", graph = ALG.adj, 
                             hyper = list(phi = list(
                                                    prior = "pc",
                                                    param = c(0.5, 2/3)) ,
                                          prec = list(
                                                    prior = "pc.prec",
                                                    param = c(0.5/0.31, 0.01))
                                          )
                             )

mod.cov <- inla(formula.cov, family = "poisson", data = full_poly@data, E = E, verbose = F, 
                control.compute = list(cpo = TRUE, dic = TRUE, graph = TRUE, waic =TRUE),
                control.predictor=list(compute=TRUE))
```


# Model evaluation

```{r}
mod.cov$summary.fixed

mod.cov$summary.hyperpar

mod.cov$dic$dic
mod.cov$dic$p.eff
mod.cov$waic$waic
mod.cov$waic$p.eff
mod.cov$mlik[1]

-sum(log(mod.cov$cpo$cpo))

fail <- mod.cov$cpo$failure
fail 
sum(mod.cov$cpo$failure >0)
```

#

```{r}
full$smr <- full$Y/full$E
full$RR <- mod.cov$summary.fitted.values[, "mean"]
full$LL <- mod.cov$summary.fitted.values[, "0.025quant"]
full$UL <- mod.cov$summary.fitted.values[, "0.975quant"]

gRR <- ggplot(full) + geom_sf(aes(fill = RR)) +
  scale_fill_gradient2(midpoint = 1, low = "blue", mid = "white", high = "red", limits = c(0.8, 1.3)) +
  theme_bw()

gLL <- ggplot(full) + geom_sf(aes(fill = LL)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red") +
  theme_bw()

gUL <- ggplot(full) + geom_sf(aes(fill = UL)) +
  scale_fill_gradient2(midpoint = 1, low = "blue", mid = "white", high = "red") +
  theme_bw()

gLL + gRR + gUL

p1 <- ggplot(full) + geom_sf(aes(fill = smr)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", 
    high = "red", limits = c(0.7,1.7)) + 
  theme_bw() +
  labs(fill = "SMR")

p2 <- ggplot(full) + geom_sf(aes(fill = RR)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", 
    high = "red", limits = c(0.8,1.3)) + 
  theme_bw() +
  labs(fill = "RR")

p1 + p2

```

