---
title: "Geostatistics"
author: "AyouB"
date: "04/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Geostatistical part

# Loading packages

```{r }
pacman::p_load(tidyverse,
               INLA,
               sp,
               sf,
               rgdal,
               rgeos,
               tmap,
               tmaptools,
               readxl,
               RColorBrewer,
               spdep,
               raster,
               viridis,
               patchwork)
```

## multi thread option 

```{r}
inla.setOption(num.threads=1)
```


# Help Functions


```{r }
logit <- function(x){log(x/(1-x))}
expit <- function(x) {exp(x)/(1+exp(x))}
```



# Data (same as areal model if already executed)

```{r}
# Import Data
data = read_xlsx("data.xlsx", sheet = "Sheet1")
data$ID <- 1:48

# Getting shapefiles
dz_mapw <- getData("GADM", country ="DZ", level = "1")
dz_map <- getData("GADM", country ="DZ", level = "0")

dz2 <- st_as_sf(dz_mapw)

# cleaning wilayas names
Wilaya <- dz2$NAME_1 %>% 
                lapply(gsub, pattern="ï", replacement = "i") %>% 
                unlist()
dz2$Wilaya <- Wilaya

# Merging data with shapefiles
full <- dz2 %>% full_join(data, by = "Wilaya")
full$CIG = full$CIG/1000

# Convert to spatial polygon again !
full_poly <- as_Spatial(full)
```



## Geostatistics

# Creation of centroids

```{r}
sf_cent <- st_centroid(full)
sp_cent <- gCentroid(as(full, "Spatial"), byid = TRUE)

sample3 <- as_Spatial(sf_cent) 

ggplot() + 
  geom_sf(data = full, fill = 'white') +
  geom_sf(data = sf_cent, color = 'red') 

ggplot() + 
  geom_sf(data = full, fill = 'white') +
  geom_sf(data = sample3 %>% st_as_sf, color = 'blue') 
```

# plot the data 

```{r}
ggplot() + 
  geom_sf(data = full, fill = 'white') +
  geom_sf(data = sf_cent, aes(col= Y2)) +
  coord_sf() +
  scale_color_gradient(low = "blue", high = "red") +
  labs(color ='Mortlaité') +
  theme_bw()

ggplot() + 
  geom_sf(data = sf_cent, aes(col= Y2)) +
  coord_sf() +
  scale_color_gradient(low = "blue", high = "red") +
  labs(color ='Mortlaité') +
  theme_bw() 
  
```

## statistics about data 

```{r}
full_poly@data %>%
  ggplot(aes(x = Y2)) +
  geom_histogram(bins = 10, fill = "blue") +
  geom_vline(xintercept = mean(data$Y), col ="red") +
  labs(
       x="Mortalité", y="Fréquence") +
  theme_bw()

full_poly@data %>%
  ggplot(aes(y = Y2)) +
  geom_boxplot(fill = "lightblue") +
  geom_hline(yintercept = mean(data$Y2), col ="red") +
  labs(y="Mortalité") + 
  theme_bw()
```


## km ?

```{r}
coo <- sample3@coords

km <- SpatialEpi::latlong2grid(coo)
km2 <- as.matrix(km)

summary(dist(km2))
```


# Mesh Creation 

```{r}
coo <- sample3@coords


mesh3 <- inla.mesh.2d(loc = coo, 
                      offset = c(5,1),
                      max.edge = c(1,3), 
                      cutoff= 0.1)
plot(mesh3)
plot(dz_map, add= T, border = "blue")
points(coo, col = "red")

```


# Other mesh specifications (don't execute !!!)

```{r}

# mesh3 <- inla.mesh.2d(loc = coo, 
#                       max.edge = c(0.1,3), 
#                       cutoff= 0.1)
# p1 <- plot(mesh3); plot(dz_map, add= T, border = "blue") ;points(coo, col = "red")
# 
# 
# mesh3 <- inla.mesh.2d(loc = coo, 
#                       max.edge = c(0.75,3), 
#                       cutoff= 0.1)
# p2<- plot(mesh3); plot(dz_map, add= T, border = "blue"); points(coo, col = "red")
# 
# 
# 
# 
# mesh3 <- inla.mesh.2d(loc = coo, 
#                       max.edge = c(1,3), 
#                       cutoff= 0.1)
# p3<- plot(mesh3); plot(dz_map, add= T, border = "blue"); points(coo, col = "red")
# 
# 
# 
# 
# mesh3 <- inla.mesh.2d(loc = coo, 
#                       max.edge = c(1.5,3), 
#                       cutoff= 0.1)
# p4 <- plot(mesh3); plot(dz_map, add= T, border = "blue"); points(coo, col = "red")
# 
# 
# 
# mesh3 <- inla.mesh.2d(loc = coo, 
#                       max.edge = c(2,3), 
#                       cutoff= 0.1)
# p5 <- plot(mesh3); plot(dz_map, add= T, border = "blue"); points(coo, col = "red")
# 
# 
# 
# mesh3 <- inla.mesh.2d(loc = coo, 
#                       max.edge = c(2.5,3), 
#                       cutoff= 0.1)
# p6 <- plot(mesh3); plot(dz_map, add= T, border = "blue"); points(coo, col = "red")
# 
# 
# mesh3 <- inla.mesh.2d(loc = coo, 
#                       offset = c(3.1),
#                       max.edge = c(1,3), 
#                       cutoff= 0.1)
# p7 <- plot(mesh3); plot(dz_map, add= T, border = "blue"); points(coo, col = "red")
# 
# 
# mesh3 <- inla.mesh.2d(loc = coo, 
#                       offset = c(4.1),
#                       max.edge = c(1,3), 
#                       cutoff= 0.1)
# p8 <- plot(mesh3); plot(dz_map, add= T, border = "blue"); points(coo, col = "red")
# 
# 
# mesh3 <- inla.mesh.2d(loc = coo, 
#                       offset = c(5.1),
#                       max.edge = c(1,3), 
#                       cutoff= 0.1)
# p9 <- plot(mesh3); plot(dz_map, add= T, border = "blue"); points(coo, col = "red")
# 
# 
# mesh3 <- inla.mesh.2d(loc =  coo,
#                       boundary = dz_map,
#                       offset = c(5,1),
#                       max.edge = c(1,3),
#                       cutoff= 0.1)
# plot(mesh3)
# plot(dz_map, add= T, border = "blue")
# points(coo, col = "red")
# 
# 
# mesh3 <- inla.mesh.2d(loc =  coo,
#                       boundary = dz_map,
#                       offset = c(5,1),
#                       max.edge = c(3,3),
#                       cutoff= 0.1)
# plot(mesh3)
# plot(dz_map, add= T, border = "blue")
# points(coo, col = "red")
# 
# 
# mee <- inla.nonconvex.hull(coo, 0.75, 0.75)
# mesh3 <- inla.mesh.2d(loc.domain =  coo,
#                       boundary = mee,
#                       max.edge = c(3,3),
#                       cutoff= 0.1)
# 
# plot(mesh3)
# points(coo, col = "red")
# 
# 
# mee <- inla.nonconvex.hull(coo, 0.9, 0.9)
# mesh3 <- inla.mesh.2d(loc.domain =  coo,
#                       boundary = mee,
#                       max.edge = c(3,3),
#                       cutoff= 0.1)
# 
# plot(mesh3)
# points(coo, col = "red")
# 
# mee <- inla.nonconvex.hull(coo, 0.6, 0.6)
# mesh3 <- inla.mesh.2d(loc.domain =  coo,
#                       boundary = mee,
#                       max.edge = c(3,3),
#                       cutoff= 0.1)
# 
# plot(mesh3)
# points(coo, col = "red")
```





# Model setting and fit

```{r}
# SPDE definition 
spde3 <- inla.spde2.matern(mesh = mesh3, alpha = 2, constr = T)

# Index set creation
index3 <- inla.spde.make.index("s", spde3$n.spde)
lengths(index3)

# Projection Matrix 
A3 <- inla.spde.make.A(mesh = mesh3, loc = coo)

# Prediction locations 
bb <- bbox(dz_map)
x <- seq(bb[1, "min"] - 1, bb[1, "max"] + 1, length.out = 100)
y <- seq(bb[2, "min"] - 1, bb[2, "max"] + 1, length.out = 100)

d <- (expand.grid(x,y,KEEP.OUT.ATTRS = F))
names(d) = c("x","y")
coordinates(d)=~x+y
d@proj4string = dz_map@proj4string 
d2 <- d[dz_map,]
plot(d2, asp = 1, type = "o")

d3 <- d2 %>% st_as_sf()

ggplot(d3) +
  geom_sf(aes(shape = "o")) + 
  theme_bw() +
  theme(legend.position = "none") 

coop <- as.matrix(data.frame(x = d2$x, y= d2$y))

# Projection Matrix for the prediction 
Ap3 <- inla.spde.make.A(mesh = mesh3, loc = coop)

# INLA Stack 
stk.e3 <- inla.stack(
  tag = "est",
  data = list(Y = full$Y2),
  A = list(1, A3),
  effects = list(data.frame(b0 = rep(1, nrow(coo)), ID = full$ID, CIG =full$CIG)
                 , s = index3)
)

stk.p3 <- inla.stack(
  tag = "pred",
  data = list(Y = NA),
  A = list(1, Ap3),
  effects = list(data.frame(b0 = rep(1, nrow(coop))), s = index3)
)

stk.full3 <- inla.stack(stk.e3, stk.p3)

# Formulas 
formula3 <- Y ~ 0 + b0 + f(s, model = spde3) 

# Model Estimation
res3 <- inla(formula3, family ="Poisson",
            data = inla.stack.data(stk.full3),
            control.predictor = list(
              compute = TRUE,
              A = inla.stack.A(stk.full3)
            ), verbose = F,
            control.compute = list(cpo = TRUE, dic = TRUE, graph = TRUE, waic =TRUE)
)
```



# Model Summary

```{r}
res3$summary.fitted.values

res3$summary.fixed

res3$summary.hyperpar
```

# Model Evaluation 

```{r}
paste("DIC")
res3$dic$dic
paste("Effective number of paramètres")
res3$dic$p.eff
paste("WAIC")
res3$waic$waic
res3$waic$p.eff
paste("Marginal likelihood")
res3$mlik
paste("log CPO")
-sum(log(res3$cpo$cpo), na.rm = TRUE)
```

# failure CPO probabilities

```{r}
fp <- res3$cpo$failure[1:48]
fp

table(fp == 0)
```


# Modified Model 

```{r}
# this procedure take a lot of time
res4 <- res3
res3 <- inla.cpo(res3)
```

# CPO plot 

```{r}
n = length(full$Y)
plot(1:n, res3$cpo$cpo[1:48], ylab="CPO",xlab = "wilayas")
abline(h = 0.01, col ="red")
```

# PIT plot

```{r}
pit <- res3$cpo$pit[1:48]
uniquant <- (1:48)/(48+1)
plot(uniquant, sort(pit), xlab="quantiles uniformes",
     ylab="valeurs PIT ordonnées")
abline(0,1)

data.frame(pit = sort(pit), uni = uniquant) %>%
  ggplot(aes(x= uni, y = pit)) + 
  geom_point()+
  geom_abline(intercept = 0, slope = 1, col = "red") + 
  labs(x= "quantiles uniformes", y = "valeurs PIT ordonnées") +
  theme_bw()

data.frame(pit = logit(sort(pit)), uni = logit(uniquant)) %>%
  ggplot(aes(x= uni, y = pit)) + 
  geom_point()+
  geom_abline(intercept = 0, slope = 1, col = "red") +
  labs(x= "quantiles uniformes", y = "valeurs PIT ordonnées") +
  ggtitle("échelle logarithmique")+
  theme_bw()
```

# Model parameters 

```{r}
spde.est3 <- inla.spde2.result(inla = res3, name = "s",
                              spde = spde3, do.transf = TRUE)


paste("minimal variance")
inla.zmarginal(spde.est3$marginals.variance.nominal[[1]])
paste("nominal range")
inla.zmarginal(spde.est3$marginals.range.nominal[[1]])

plot(spde.est3[["marginals.range.nominal"]][[1]], type = "l")
plot(spde.est3[["marginals.variance.nominal"]][[1]], type = "l")
plot(spde.est3[["marginals.log.kappa"]][[1]], type = "l")
plot(spde.est3[["marginals.log.tau"]][[1]], type = "l")
```


## plot model (change the true value to get different plots)

```{r}
plot(res3,
     plot.fixed.effects = FALSE,
     plot.lincomb = FALSE,
     plot.random.effects = FALSE,
     plot.hyperparameters = FALSE,
     plot.predictor = FALSE,
     plot.q = FALSE,
     plot.cpo = TRUE
)
```


## RMSE on train data

```{r}
index.est = inla.stack.index(stk.full3,"est")$data
lp.mean.val = res3$summary.linear.predictor[index.est,"mean"]
e <- exp(lp.mean.val)

paste("corrélation")
cor(e, full_poly@data$Y2)

rest = full_poly@data$Y2 - e

paste0("RMSE")
sqrt(mean(rest^2))

data.frame(Y = full_poly@data$Y2, e = e) %>%
  ggplot(aes(x = Y, y = e)) + 
  geom_point() +
  labs(x = "Observation", y = "Estimation") +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  theme_bw()
  

plot(full_poly@data$Y, e, xlab="Observation", ylab="prédiction")
```



# Plot predictions

```{r}
index_3 <- inla.stack.index(stk.full3, tag = "pred")$data

pred_mean3 <- exp(res3$summary.linear.predictor[index_3, "mean"])
pred_ll3 <- exp(res3$summary.linear.predictor[index_3, "0.025quant"])
pred_ul3 <- exp(res3$summary.linear.predictor[index_3, "0.975quant"])

dpm3 <- rbind(

  data.frame(
    east = coop[, 1], north = coop[, 2],
    value = pred_ll3, variable = "borne inférieure"
  ),
  data.frame(
    east = coop[, 1], north = coop[, 2],
    value = pred_mean3, variable = "prédiction moyenne"
  ),
  data.frame(
    east = coop[, 1], north = coop[, 2],
    value = pred_ul3, variable = "borne supérieure"
  )
)
dpm3$variable <- as.factor(dpm3$variable)

dpm3 %>%
  filter(variable == "prédiction moyenne") %>%
  ggplot() + geom_tile(aes(east, north, fill = value)) +
  #facet_wrap(~variable, nrow = 3) +
  coord_fixed(ratio = 1) +
  scale_fill_gradient(
    name = "Mortalité",
    low = "blue", high = "red"
  ) +
  theme_bw()

dpm3 %>%
  #filter(variable != "prédiction moyenne") %>%
  ggplot() + geom_tile(aes(east, north, fill = value)) +
  facet_wrap(~variable, nrow = 1) +
  coord_fixed(ratio = 1) +
  scale_fill_gradient(
    name = "Mortalité",
    low = "blue", high = "orange"
  ) +
  theme_bw()

dpm3 %>%
  group_by(variable) %>%
  summarize(v = min(value))
```

# Spatial Field projection

```{r}
rang3 <- apply(mesh3$loc[, c(1, 2)], 2, range)
proj3 <- inla.mesh.projector(mesh3,
                             xlim = rang3[, 1], ylim = rang3[, 2],
                             dims = c(1000, 1000))

mean_s3 <- inla.mesh.project(proj3, res3$summary.random$s$mean)
sd_s3 <- inla.mesh.project(proj3, res3$summary.random$s$sd)

df3 <- expand.grid(x = proj3$x, y = proj3$y)
df3$mean <- as.vector(mean_s3)
df3$sd <- as.vector(sd_s3)



gmean3 <- ggplot(df3, aes(x = x, y = y, fill = mean)) +
  geom_raster() + 
  scale_fill_viridis(na.value = "transparent") +
  labs(x = "east", y = "north" , fill = "moyenne") +
  coord_fixed(ratio = 1) + theme_bw()

gsd3 <- ggplot(df3, aes(x = x, y = y, fill = sd)) +
  geom_raster() +
  scale_fill_viridis(na.value = "transparent") +
  labs(x = "east", y = "north" , fill = "écart-type") +
  coord_fixed(ratio = 1) + theme_bw()

gmean3 + gsd3
```

# same plots but with the study region (takes time)

```{r}
coordinates(df3) =~x+y
df3@proj4string = dz_map@proj4string
df3 <- as.data.frame(df3[dz_map,])



gmean3 <- ggplot(df3, aes(x = x, y = y, fill = mean)) +
  geom_raster() +
  scale_fill_viridis(na.value = "transparent") +
  labs(x = "east", y = "north" , fill = "moyenne") +
  coord_sf() + theme_bw()

gsd3 <- ggplot(df3, aes(x = x, y = y, fill = sd)) +
  geom_raster() +
  scale_fill_viridis(na.value = "transparent",begin = 0.17, end =0.8) +
  labs(x = "east", y = "north" , fill = "écart-type") +
  coord_sf() + theme_bw()


gmean3 + gsd3
gsd3
gmean3
```


## PC prior model (another model specification)


```{r}
coo <- sample3@coords
## summary(dist(coo))


mesh3 <- inla.mesh.2d(loc = coo, 
                      offset = c(5,1),
                      max.edge = c(1,3), 
                      cutoff= 0.1)
plot(mesh3)
plot(dz_map, add= T, border = "blue")
points(coo, col = "red")


spde3 = inla.spde2.pcmatern(mesh3,
                                   prior.range = c(5, 0.5),
                                   prior.sigma = c(10, 0.05))

index3 <- inla.spde.make.index("s", spde3$n.spde)
lengths(index3)

A3 <- inla.spde.make.A(mesh = mesh3, loc = coo)

bb <- bbox(dz_map)
x <- seq(bb[1, "min"] - 1, bb[1, "max"] + 1, length.out = 100)
y <- seq(bb[2, "min"] - 1, bb[2, "max"] + 1, length.out = 100)

d <- (expand.grid(x,y,KEEP.OUT.ATTRS = F))
names(d) = c("x","y")
coordinates(d)=~x+y
d@proj4string = dz_map@proj4string 
d2 <- d[dz_map,]
plot(d2, asp = 1, type = "o")

d3 <- d2 %>% st_as_sf()

ggplot(d3) +
  geom_sf(aes(shape = "o")) + 
  theme_bw() +
  theme(legend.position = "none") 

coop <- as.matrix(data.frame(x = d2$x, y= d2$y))

Ap3 <- inla.spde.make.A(mesh = mesh3, loc = coop)

stk.e3 <- inla.stack(
  tag = "est",
  data = list(Y = full$Y2),
  A = list(1, A3),
  effects = list(data.frame(b0 = rep(1, nrow(coo)), ID = full$ID, CIG =full$CIG)
                 , s = index3)
)

stk.p3 <- inla.stack(
  tag = "pred",
  data = list(Y = NA),
  A = list(1, Ap3),
  effects = list(data.frame(b0 = rep(1, nrow(coop))), s = index3)
)

stk.full3 <- inla.stack(stk.e3, stk.p3)

formula3 <- Y ~ 0 + b0  + f(s, model = spde3)

pc3 <- inla(formula3, family ="Poisson",
            data = inla.stack.data(stk.full3),
            control.predictor = list(
              compute = TRUE,
              A = inla.stack.A(stk.full3)
            ), verbose = T,
            control.compute = list(cpo = TRUE, dic = TRUE, graph = TRUE, waic =TRUE)
)
```

## Model with PC evaluation 

```{r}
pc3$summary.fitted.values

pc3$summary.fixed

pc3$summary.hyperpar


paste("DIC")
pc3$dic$dic
paste("Effective number of paramètres")
pc3$dic$p.eff
paste("WAIC")
pc3$waic$waic
pc3$waic$p.eff
paste("Marginal likelihood")
pc3$mlik
paste("log CPO")
-sum(log(pc3$cpo$cpo), na.rm = TRUE)



```

## CPO correction 

```{r}
pc4 <- pc3
pc3 <- inla.cpo(pc3)
```


# Plot predictions

```{r}
index_3 <- inla.stack.index(stk.full3, tag = "pred")$data

pred_mean3 <- exp(pc3$summary.linear.predictor[index_3, "mean"])
pred_ll3 <- exp(pc3$summary.linear.predictor[index_3, "0.025quant"])
pred_ul3 <- exp(pc3$summary.linear.predictor[index_3, "0.975quant"])

dpm3 <- rbind(

  data.frame(
    east = coop[, 1], north = coop[, 2],
    value = pred_ll3, variable = "borne inférieure"
  ),
  data.frame(
    east = coop[, 1], north = coop[, 2],
    value = pred_mean3, variable = "prédiction moyenne"
  ),
  data.frame(
    east = coop[, 1], north = coop[, 2],
    value = pred_ul3, variable = "borne supérieure"
  )
)
dpm3$variable <- as.factor(dpm3$variable)

dpm3 %>%
  filter(variable == "prédiction moyenne") %>%
  ggplot() + geom_tile(aes(east, north, fill = value)) +
  #facet_wrap(~variable, nrow = 3) +
  coord_fixed(ratio = 1) +
  scale_fill_gradient(
    name = "Mortalité",
    low = "blue", high = "red"
  ) +
  theme_bw()

dpm3 %>%
  filter(variable != "prédiction moyenne") %>%
  ggplot() + geom_tile(aes(east, north, fill = value)) +
  facet_wrap(~variable, nrow = 1) +
  coord_fixed(ratio = 1) +
  scale_fill_gradient(
    name = "Mortalité",
    low = "blue", high = "orange"
  ) +
  theme_bw()

dpm3 %>%
  group_by(variable) %>%
  summarize(v = min(value))




rang3 <- apply(mesh3$loc[, c(1, 2)], 2, range)
proj3 <- inla.mesh.projector(mesh3,
                             xlim = rang3[, 1], ylim = rang3[, 2],
                             dims = c(1000, 1000))

mean_s3 <- inla.mesh.project(proj3, pc3$summary.random$s$mean)
sd_s3 <- inla.mesh.project(proj3, pc3$summary.random$s$sd)

df3 <- expand.grid(x = proj3$x, y = proj3$y)
df3$mean <- as.vector(mean_s3)
df3$sd <- as.vector(sd_s3)



gmean3 <- ggplot(df3, aes(x = x, y = y, fill = mean)) +
  geom_raster() + 
  scale_fill_viridis(na.value = "transparent") +
  labs(x = "east", y = "north" , fill = "moyenne") +
  coord_fixed(ratio = 1) + theme_bw()

gsd3 <- ggplot(df3, aes(x = x, y = y, fill = sd)) +
  geom_raster() +
  scale_fill_viridis(na.value = "transparent") +
  labs(x = "east", y = "north" , fill = "écart-type") +
  coord_fixed(ratio = 1) + theme_bw()

gmean3 + gsd3


coordinates(df3) =~x+y
df3@proj4string = dz_map@proj4string
df3 <- as.data.frame(df3[dz_map,])


gmean3 <- ggplot(df3, aes(x = x, y = y, fill = mean)) +
  geom_raster() +
  scale_fill_viridis(na.value = "transparent") +
  labs(x = "east", y = "north" , fill = "moyenne") +
  coord_sf() + theme_bw()

gsd3 <- ggplot(df3, aes(x = x, y = y, fill = sd)) +
  geom_raster() +
  scale_fill_viridis(na.value = "transparent",begin = 0.17, end =0.8) +
  labs(x = "east", y = "north" , fill = "écart-type") +
  coord_sf() + theme_bw()


gmean3 + gsd3
gsd3
gmean3

```
